Enzo+Moray: Radiation Hydrodynamics Adaptive Mesh Refinement Simulations with Adaptive Ray Tracing

### 2010-08-31 ms.tex
Adding results and figures for Test 10 and Test 12 (hydro).  Adding weak scaling figure and description.

  We describe a photon-conserving numerical method that solves the
  radiative transfer equation, using a spatially-adaptive ray tracing
  scheme, and its parallel implementation into the adaptive mesh
  refinement (AMR) code, Enzo.  Coupling the solver with the energy
  equation and non-equilibrium chemistry network, our radiation
  hydrodynamics framework can be utilized to study a broad range of
  astrophysical problems, such as stellar and black hole (BH)
  feedback.  Inaccuracies can arise from large timesteps and poor
  sampling, therefore we devised an adaptive timestepping scheme and a
  fast approximation of the optically-thin radiation field with
  multiple sources.  We test the method with several radiative
  transfer and radiation hydrodynamics tests that are given in
  \citet{RT06, Iliev09}.  We further test our method with more
  dynamical situations, for example, the propagation of an ionization
  front through a Rayleigh-Taylor instability, time-varying
  luminosities, and collimated radiation.  This method linearly scales
  with the number of point sources and number of grid cells.  To
  combat this, we use a novel method of merging rays at large radius,
  using a tree method, which we briefly describe.  Our implementation
  is scalable to O(10$^3$) processors on distributed and shared memory
  machines and can include radiation pressure and secondary
  ionizations from X-ray radiation.  It is included in the newest
  public release of Enzo.

@@

Fortunately problems can be simplified by exploiting symmetries and
special cases.  The first case was considered by \citet{Stroemgren39}
for a radiation source ionizing a static uniform neutral medium, where
recombinations balnace photo-ionizations.  He derived a radius of an
\ion{H}{2} region,
%
\begin{equation}
  \label{eqn:Rstr}
  R_s = \left(\frac{3\dot{N}_\gamma}{4\pi \alpha_{\rm B} n_H^2}\right)^{1/3},
\end{equation}
where $\dot{N}_\gamma$ is the ionizing photon luminosity, $\alpha_{\rm
  B}$ is the recombination rate, and $n_H$ is the ambient hydrogen
number density.  Furthermore he found that the delineation between the
neutral and ionized medium to be approximately the mean free path of
the ionizing radiation.  His seminal work was expanded upon by
\citet{Spitzer48, Spitzer49, Spitzer54} and \citet{Spitzer50}, who
showed that the ionizing radiation heated the medium to $T \sim 10^4$
K.  If the density is equal on both sides of the ionization front,
then this over-pressurized region would expand and drive a shock
outwards \citep[e.g.][]{Oort54, Schatzman55}.  These early works
provided the basis for the modern topic of radiation hydrodynamics.  A
decade later, the first radiation hydrodynamical numerical models of
\ion{H}{2} regions in spherical symmetry and plane-parallel ionization
fronts were developed \citep[e.g.][]{Mathews65, Lasker66,
  Hjellming66}.  They described the expansion of the ionization front
and the evolution of its associated shock wave that carries most of
the gas away from the source.  At the same time, theoretical models of
ionization fronts matured and were classified by \citet{Kahn54} and
\citet{Axford61} as either R-type (rare) or D-type (dense).  In R-type
fronts, the ionized gas density is higher than the neutral gas
density, and in D-type fronts, the opposite is true.  R-type fronts
travel supersonically with respect to the neutral gas, whereas D-type
front are subsonic.  Furthermore ``weak'' and ``strong'' R-type fronts
move supersonically and subsonically with respect to the ionized gas,
respectively.  The same terminology conversely applies to D-type
fronts.  ``Critical'' fronts are defined as moving exactly at the
sound speed.  These works established the evolutionary track of an
expanding \ion{H}{2} illuminated by a massive star in a uniform
medium:
%
\begin{enumerate}
\item Weak R-type---When the star (gradually) starts to shine, the
  ionization front will move supersonically through the ambient
  medium.  The gas is heated and ionized, but otherwise left
  undisturbed.  This stage continues until $r \sim 0.02R_s$.
\item Critical R-type---As the ionization front moves outwards, it
  begins to slow because of the geometric dilution of the radiation.
  It becomes a critical R-type front, which is equivalent to an
  isothermal shock in the neutral gas.
\item Strong and weak D-type---The front continues to slow, becoming a
  strong D-type front, and then a critical D-type front.  From this
  point forward, it is moving subsonically with respect to the ionized
  gas, i.e. a weak D-type front.  Thus sound waves can travel across
  the ionization front and form a shock.  The ionization front
  detaches from the shock, putting the shock ahead of the ionization
  front.
\item Expansion phase---After the shock forms, the \ion{H}{2} region
  starts to expand, lowering the interior density and thus the
  recombination rates.  This increases the number of photons available
  for ionizing the gas.  The sphere expands until it reaches pressure
  equilibrium with the ambient medium at $r \sim 5R_s$.
\end{enumerate}
%

In the 1970's and 1980's, algorithmic and computational advances
allowed numerical models to be expanded to two dimensions, mainly
using axi-symmetric to simplify the problem
\citep[e.g.][]{Bodenheimer79, Sandford82, Yorke83}.  One topic that was
studied extensively were champange flows.  Here the source is embedded
in an overdense region, and the \ion{H}{2} region escapes from this
region in one direction.  The interface between the ambient and dense
medium was usually set up to be a constant pressure boundary.  When
the ionization front passes this boundary, the dense, ionized gas is
orders of magnitude out of pressure equilibrium as the temperatures on
both sides of initial boundary are within a factor of a few.  In
response, the gas is accelerated outwards in this direction, creating
a fan-shaped outflow.

Only in the past 15 years, computational resources have become large
enough, along with further algorithmic advances, to cope with the
requirements of three-dimensional calculations.  There are two popular
methods to solve the radiative transfer equation in three-dimensions:
%
\begin{itemize}
\item Moment methods---The angular moments of the radiation field can
  describe its angular structure, which are related to the energy
  energy, flux, and radiation pressure \citep{Auer70, Norman98}.
  These have been implemented in conjunction with short
  characteristics \citep[2D]{Stone92_RHD}, with long characteristics
  \citep{Finlator09}, with a variable Eddington tensor in the
  optically thin limit \citep[OTVET;][]{Gnedin01_OTVET, Petkova09},
  and with an M1 closer relation \citep{Gonzalez07, Aubert08}.  Moment
  methods have the advantage of being fast, being independent of the
  number of radiation point sources.  However, they are diffusive and
  result in incorrect shadows in some situations.
\item Ray tracing---Radiation can be propagated along rays that extend
  through the compuational grid \citep[e.g.][]{Razoumov99, Abel99_RT,
    Ciardi01, Sokasian01, Whalen06, Rijkhorst06, Mellema06, Alvarez06,
    Trac07, Krumholz07_ART, Paardekooper10} or particle set
  \citep[e.g.][]{Susa06, Johnson07, Pawlik08, Pawlik10, Altay08,
    Hasegawa09}.  In general, these methods are very accurate but
  computationally expensive because the radiation field must be well
  sampled by the rays with respect to the spatial resolution of the
  grid or particles.
\end{itemize}
%
Until the mid-2000's the vast majority of the three-dimensional
calculations were performed with static density distributions.  One
example is calculating cosmological reionization by post-processing of
density fields from N-body simulations \citep{Ciardi01, Sokasian01,
  McQuinn07, Iliev06, Iliev07}.  Any hydrodynamical response to the
radiation field was thus ignored.  Several radiative transfer codes
were compared in four purely radiative transfer tests in
\citet[hereafter RT06]{RT06}.  Only recently has the radiative
transfer equation been coupled to the hydrodynamics in
three-dimensions.  In the second comparision paper \citep[hereafter
  RT09]{Iliev09}, results from these radiation hydrodynamics codes
were compared.  Even more rare are ones that couple it with
magneto-hydrodynamics \citep[e.g.][]{Krumholz07_ART}.  The tests in
RT06 and RT09 were kept relatively simple to ease the comparison.

@@

Radiation trnasport is a well-studied topic, and we begin by
describing our approach in solving the radiative transfer equation,
which in comoving coordinates \citep{Gnedin97} is
%
\begin{equation}
  \label{eqn:rteqn}
  \frac{1}{c} \; \frac{\partial I_\nu}{\partial t} + 
  \frac{\hat{n} \cdot \nabla I_\nu}{\bar{a}} -
  \frac{H}{c} \; \left( \nu \frac{\partial I_\nu}{\partial \nu} -
  3 I_\nu \right) = -\kappa_\nu I_\nu + j_\nu .
\end{equation}
%
Here $I_\nu \equiv I(\nu, \mathbf{x}, \Omega, t)$ is the radiation
specific intensity in units of energy per time $t$ per solid angle per
unit area per frequency $\nu$.  $H = \dot{a}/a$ is the Hubble
constant, where $a$ is the scale factor.  $\bar{a} = a/a_{em}$ is the
ratio of scale factors at the current time and time of emission.  The
second term represents the propagation of radiation, where the factor
$1/a$ accounts for cosmic expansion.  The third term describes both
the cosmological redshift and dilution of radiation.  On the right
hand side, the first term considers the absorption coefficient
$\kappa_\nu \equiv \kappa_\nu(\mathbf{x},\nu,t)$, and the second term
$j_\nu \equiv j_\nu(\mathbf{x},\nu,t)$ is the emission coefficient
that includes any point sources of radiation or diffuse radiation.

Solving this equation is difficult because of its high dimensionality;
however, we can make some appropriate approximations to reduce its
complexity in order to include radiation transport in numerical
calculations.  Typically timesteps in dynamic calculations are small
enough so that $\Delta a/a \ll 1$, therefore $\bar{a} = 1$ in any
given timestep, reducing the second term to $\hat{n} \partial
I_\nu/\partial \mathbf{x}$.  To determine the importance of the third
term, we evaluate the ratio of the third term ot the second term.
This is $HL/c$, where $L$ is the simulation box length.  If this ratio
is $\ll 1$, we can ignore the third term.  For example at $z=5$, this
ratio is 0.1 when $L = c/H(z=5)$ = 53 proper Mpc.  In large boxes
where the light crossing time is comparable to the Hubble time, then
it could be important to consider cosmological redshifting and
dilution of the radiation, which we will describe later in
\S\ref{sec:ART}.  Thus equation (\ref{eqn:rteqn}) reduces to the
non-cosmological form in this local approximation,
%
\begin{equation}
  \frac{1}{c} \frac{\partial I_\nu}{\partial t} + 
  \hat{n} \frac{\partial I_\nu}{\partial \mathbf{x}} =
  -\kappa_\nu I_\nu + j_\nu .
\end{equation}
%
We choose to represent the source term $j_\nu$ as point sources of
radiation (e.g. stars, quasars) that emit radial rays that are
propaged along the direction $\hat{n}$.  Next we describe this
discretization and its contribution to the radiation field.

@@

The radiation field is calculated by integrating equation
\ref{eqn:rtray} along each ray, which is done by considering the
discretization of the ray into segments.  In the following section, we
assume the rays are monochromatic.  For convenience, we express the
integration in terms of optical depth $\tau = \int \kappa(r,t) \;
dr$, and for a ray segment,
%
\begin{equation}
  \label{eqn:dtau}
  d\tau = \sigma_{\rm abs}(\nu) n_{\rm abs} dr.
\end{equation}
Here $\sigma_{\rm abs}$ and $n_{\rm abs}$ are the cross section and
number density of the absorbing medium, respectively.  We use the
cell-centered density in our calculations but have experimented with
trilinearly interpolated densities \citep[see][]{Mellema06} without
producing considerable artifacts.  Equation (\ref{eqn:rtray}) has a
simple exponential analytic solution, and the photon flux of a ray is
reduced by
%
\begin{equation}
  \label{eqn:flux}
  dP = P \times (1 - e^{-\tau})
\end{equation}
as it crosses a cell.  We equate the photo-ionization rate to the
absorption rate, resulting in photon conservation \citep{Abel99_RT,
  Mellema06}.  Thus the photo-ionization $k_{\rm ph}$ and
photo-heating $\Gamma_{\rm ph}$ rates associated with a single ray are
%
\begin{equation}
  \label{eqn:kph}
  k_{\rm ph} = \frac{P (1 - e^{-\tau})}{n_{\rm abs} \; V_{\rm cell} \; dt_P},
\end{equation}
\begin{equation}
  \label{eqn:gamma}
  \Gamma_{\rm ph} = k_{\rm ph} \; (E_{\rm ph} - E_i),
\end{equation}
where $V_{\rm cell}$ is the cell volume, $E_{\rm ph}$ is the photon
energy, and $E_i$ is the ionization energy of the absorbing material.
In each cell, the photo-ionization and photo-heating rates from each
ray in the calculation are summed, and after the ray tracing is
complete, these rates can be used to update the ionization state and
energy of the cells.  Considering a system with only hydrogen
photo-ionizations and radiative recombinations, these changes are very
straightforward and is useful for illustrative purposes.  The change
in neutral hydrogen is
%
\begin{equation}
\frac{dn_{\rm H}}{dt} = \alpha_B n_e n_p - C_{\rm H} n_e n_{\rm
  H} - k_{\rm ph},
\end{equation}
where $\alpha_B = 2.59 \times 10^{-13} \mathrm{cm}^3 \mathrm{s}^{-1}$
is the recombination coefficent at 10$^4$ K in the Case B on-the-spot
approximation, in which all recombinations are locally reabsorbed,
\citep{Spitzer78}, and $C_{\rm H}$ is the collisional ionization rate.
However for more accurate solutions in calculations that consider
several chemical species, the photo-ionization rates are better
utilized in solvers that consider chemical networks
\citep[e.g.][]{Abel97}.

@@

Other radiative processes can also be important in some situations,
such as attentuation of radiation in the Lyman-Werner bands, secondary
ionizations from X-ray radiation, Compton heating of from scattered
photons, and radiation pressure.

@@

Molecular hydrogen can absorb photons in the Lyman-Werner bands that
are composed of 76 absoprtion lines ranging from 10.2 to 13.6 eV.
Each of these spectral lines can be modelled with a typical
exponential attenuation equation \citep{Ricotti01}, but
\citet{Draine96} showed that this self-shielding is well modeled with
the following relation to total \hh~column density
%
\begin{equation}
  \label{eqn:lwband}
  f_{\rm shield}(N_{\rm H2}) = \left\{ \begin{array}{l@{\quad}l}
      1 & (N_{\rm H2} \le 10^{14} \; \mathrm{cm}^{-2})\\
      (N_{\rm H2}/10^{14} \; \mathrm{cm}^{-2})^{-0.75} & (N_{\rm H2} >
        10^{14} \; \mathrm{cm}^{-2})
    \end{array} \right. .
\end{equation}
To incorporate this shielding function into the ray tracer, we store
the total \hh~column density and calculate the \hh~dissociation rate
by summing the contribution of all rays,
%
\begin{equation}
  \label{eqn:lwRT}
  k_{\rm diss} = \sum_{\rm rays} \frac{P \; \sigma_{\rm LW} \;
    \Omega_{\rm ray} \; r^2 \; dr}{A_{\rm cell} \; dV \; dt_P},
\end{equation}
where $\sigma_{\rm LW} = 3.71 \times 10^{18} \mathrm{cm}^2$ is the
effective cross-section of \hh~\citep{Abel97}.  To account for any
absorption in each photon package, we attentuate the photon number
flux by
%
\begin{equation}
  \label{eqn:LWdP}
  dP = P [f_{\rm shield}(N_{\rm H2} + dN_{\rm H2}) - f_{\rm shield}(N_{\rm H2})],
\end{equation}
where $dN_{\rm H2}$ is the \hh~column density in the current cell.

@@

Consider the solid angle $\Omega_{\rm ray}$ and photon flux $P$
associated with a single ray, and assume the flux is constant across
$\Omega_{\rm ray}$.  There exists a discrenpancy between the geometry
cell face and HEALPix pixel when the pixel does not cover the entire
cell face, which is illustrated in Figure \ref{fig:covering}.  This
mismatch causes non-spherical artifacts and is most apparent in the
optically thin case, where the area of the pixel is dominant over $(1
- e^\tau)$ when calculating $k_{\rm ph}$.  One can avoid these
artifacts by increasing the sampling $\Phi_c$ to high values,
e.g. $>10$, but we have formulated a simple geometric correction to
the calculation of the radiation field.  This correction is not unique
to the HEALPix formalism but can be applied to any type of
pixelization.

@@

When the medium is optically thin to the radiation, radiation is only
attentuated by geometric dilution in the local approximation to
Eq. (\ref{eqn:rteqn}), i.e. the inverse square law.  We can minimize
the computational work of ray tracing in the optically thin regime by
utilizing this fact.  Here we track the total column density $N_{\rm
  abs}$ and the equivalent total optical depth $\tau$ traversed by the
photon.  If $\tau < \tau_{\rm thin} \sim 0.1$ after the ray exits the
cell, we calculate the photo-ionization and photo-heating rates
directly from the incoming ray instead of the luminosity of the
source.  This should only be evaluated once per cell per radiation
source.
\begin{equation}
  \label{eqn:optthin}
  k_{\rm ph} = \frac{\sigma_{\rm abs} \; P}{dt_P \; \theta_{\rm pix}}
  \; \frac{r_{\rm cell}}{r_{\rm ray}}.
\end{equation}
Note that the photon number $P$ in the ray has already been
geometrically diluted by ray splitting.  Here $r_{\rm cell}$ and
$r_{\rm ray}$ are the radii from the radiation source to the cell
center and where the ray exits the cell.  Thus the last factor
corrects the flux to a value appropriate for the cell center.  The
photo-heating ray is calculated in the same manner as the general
case, $\Gamma_{\rm ph} = k_{\rm ph} (E_{\rm ph} - E_i)$.  No photons
are removed from the ray.  With this method, we only require one ray
travel through each cell where the gas is optically thin, thus
reducing the computational expense.

@@

\enzo~is a parallel block-structured AMR \citep{BergerAMR} code that
is publicly available \citep{BryanNorman1997, OShea2004}.  In this
section, we describe our parallel implementation of the adaptive ray
tracing method into \enzo.  First we explain the programming design of
handling the ``photon packages'' that are traced along the adaptive
rays.  We use the terms photon packages and rays interchangably.  Next
we focus on the details of the radiation hydrodynamics and then the
importance of correct timestepping.  Last we give our parallelization
strategy of tracing rays through an AMR hierarchy.  This
implementation is included in the \texttt{v2.0} public version of
\enzo.

@@

\subsubsection{Time averaged quantites within a timestep}

>>>>

  We describe a photon-conserving numerical method that solves the
  radiative transfer equation, using a spatially-adaptive ray tracing
  scheme, and its parallel implementation into the adaptive mesh
  refinement (AMR) code, Enzo.  Coupling the solver with the energy
  equation and non-equilibrium chemistry network, our radiation
  hydrodynamics framework can be utilized to study a broad range of
  astrophysical problems, such as stellar and black hole (BH)
  feedback.  Inaccuracies can arise from large timesteps and poor
  sampling, therefore we devised an adaptive time-stepping scheme and a
  fast approximation of the optically-thin radiation field with
  multiple sources.  We test the method with several radiative
  transfer and radiation hydrodynamics tests that are given in
  \citet{RT06, Iliev09}.  We further test our method with more
  dynamical situations, for example, the propagation of an ionization
  front through a Rayleigh-Taylor instability, time-varying
  luminosities, and collimated radiation.  This method linearly scales
  with the number of point sources and number of grid cells.  To
  combat this, we use a novel method of merging rays at large radius,
  using a tree method, which we briefly describe.  Our implementation
  is scalable to O(10$^3$) processors on distributed and shared memory
  machines and can include radiation pressure and secondary
  ionizations from X-ray radiation.  It is included in the newest
  public release of Enzo.

@@

Fortunately problems can be simplified by exploiting symmetries and
special cases.  The first case was considered by \citet{Stroemgren39}
for a radiation source ionizing a static uniform neutral medium, where
recombinations balance photo-ionizations.  He derived a radius of an
\ion{H}{2} region,
%
\begin{equation}
  \label{eqn:Rstr}
  R_s = \left(\frac{3\dot{N}_\gamma}{4\pi \alpha_{\rm B} n_H^2}\right)^{1/3},
\end{equation}
where $\dot{N}_\gamma$ is the ionizing photon luminosity, $\alpha_{\rm
  B}$ is the recombination rate, and $n_H$ is the ambient hydrogen
number density.  Furthermore he found that the delineation between the
neutral and ionized medium to be approximately the mean free path of
the ionizing radiation.  His seminal work was expanded upon by
\citet{Spitzer48, Spitzer49, Spitzer54} and \citet{Spitzer50}, who
showed that the ionizing radiation heated the medium to $T \sim 10^4$
K.  If the density is equal on both sides of the ionization front,
then this over-pressurized region would expand and drive a shock
outwards \citep[e.g.][]{Oort54, Schatzman55}.  These early works
provided the basis for the modern topic of radiation hydrodynamics.  A
decade later, the first radiation hydrodynamical numerical models of
\ion{H}{2} regions in spherical symmetry and plane-parallel ionization
fronts were developed \citep[e.g.][]{Mathews65, Lasker66,
  Hjellming66}.  They described the expansion of the ionization front
and the evolution of its associated shock wave that carries most of
the gas away from the source.  At the same time, theoretical models of
ionization fronts matured and were classified by \citet{Kahn54} and
\citet{Axford61} as either R-type (rare) or D-type (dense).  In R-type
fronts, the ionized gas density is higher than the neutral gas
density, and in D-type fronts, the opposite is true.  R-type fronts
travel supersonically with respect to the neutral gas, whereas D-type
front are subsonic.  Furthermore ``weak'' and ``strong'' R-type fronts
move supersonically and subsonically with respect to the ionized gas,
respectively.  The same terminology conversely applies to D-type
fronts.  ``Critical'' fronts are defined as moving exactly at the
sound speed.  These works established the evolutionary track of an
expanding \ion{H}{2} illuminated by a massive star in a uniform
medium:
%
\begin{enumerate}
\item Weak R-type---When the star (gradually) starts to shine, the
  ionization front will move supersonically through the ambient
  medium.  The gas is heated and ionized, but otherwise left
  undisturbed.  This stage continues until $r \sim 0.02R_s$.
\item Critical R-type---As the ionization front moves outwards, it
  begins to slow because of the geometric dilution of the radiation.
  It becomes a critical R-type front, which is equivalent to an
  isothermal shock in the neutral gas.
\item Strong and weak D-type---The front continues to slow, becoming a
  strong D-type front, and then a critical D-type front.  From this
  point forward, it is moving subsonically with respect to the ionized
  gas, i.e. a weak D-type front.  Thus sound waves can travel across
  the ionization front and form a shock.  The ionization front
  detaches from the shock, putting the shock ahead of the ionization
  front.
\item Expansion phase---After the shock forms, the \ion{H}{2} region
  starts to expand, lowering the interior density and thus the
  recombination rates.  This increases the number of photons available
  for ionizing the gas.  The sphere expands until it reaches pressure
  equilibrium with the ambient medium at $r \sim 5R_s$.
\end{enumerate}
%

In the 1970's and 1980's, algorithmic and computational advances
allowed numerical models to be expanded to two dimensions, mainly
using axi-symmetric to simplify the problem
\citep[e.g.][]{Bodenheimer79, Sandford82, Yorke83}.  One topic that was
studied extensively were champagne flows.  Here the source is embedded
in an overdense region, and the \ion{H}{2} region escapes from this
region in one direction.  The interface between the ambient and dense
medium was usually set up to be a constant pressure boundary.  When
the ionization front passes this boundary, the dense, ionized gas is
orders of magnitude out of pressure equilibrium as the temperatures on
both sides of initial boundary are within a factor of a few.  In
response, the gas is accelerated outwards in this direction, creating
a fan-shaped outflow.

Only in the past 15 years, computational resources have become large
enough, along with further algorithmic advances, to cope with the
requirements of three-dimensional calculations.  There are two popular
methods to solve the radiative transfer equation in three-dimensions:
%
\begin{itemize}
\item Moment methods---The angular moments of the radiation field can
  describe its angular structure, which are related to the energy
  energy, flux, and radiation pressure \citep{Auer70, Norman98}.
  These have been implemented in conjunction with short
  characteristics \citep[2D]{Stone92_RHD}, with long characteristics
  \citep{Finlator09}, with a variable Eddington tensor in the
  optically thin limit \citep[OTVET;][]{Gnedin01_OTVET, Petkova09},
  and with an M1 closer relation \citep{Gonzalez07, Aubert08}.  Moment
  methods have the advantage of being fast, being independent of the
  number of radiation point sources.  However, they are diffusive and
  result in incorrect shadows in some situations.
\item Ray tracing---Radiation can be propagated along rays that extend
  through the computational grid \citep[e.g.][]{Razoumov99, Abel99_RT,
    Ciardi01, Sokasian01, Whalen06, Rijkhorst06, Mellema06, Alvarez06,
    Trac07, Krumholz07_ART, Paardekooper10} or particle set
  \citep[e.g.][]{Susa06, Johnson07, Pawlik08, Pawlik10, Altay08,
    Hasegawa09}.  In general, these methods are very accurate but
  computationally expensive because the radiation field must be well
  sampled by the rays with respect to the spatial resolution of the
  grid or particles.
\end{itemize}
%
Until the mid-2000's the vast majority of the three-dimensional
calculations were performed with static density distributions.  One
example is calculating cosmological reionization by post-processing of
density fields from N-body simulations \citep{Ciardi01, Sokasian01,
  McQuinn07, Iliev06, Iliev07}.  Any hydrodynamical response to the
radiation field was thus ignored.  Several radiative transfer codes
were compared in four purely radiative transfer tests in
\citet[hereafter RT06]{RT06}.  Only recently has the radiative
transfer equation been coupled to the hydrodynamics in
three-dimensions.  In the second comparison paper \citep[hereafter
  RT09]{Iliev09}, results from these radiation hydrodynamics codes
were compared.  Even more rare are ones that couple it with
magneto-hydrodynamics \citep[e.g.][]{Krumholz07_ART}.  The tests in
RT06 and RT09 were kept relatively simple to ease the comparison.

@@

Radiation transport is a well-studied topic, and we begin by
describing our approach in solving the radiative transfer equation,
which in comoving coordinates \citep{Gnedin97} is
%
\begin{equation}
  \label{eqn:rteqn}
  \frac{1}{c} \; \frac{\partial I_\nu}{\partial t} + 
  \frac{\hat{n} \cdot \nabla I_\nu}{\bar{a}} -
  \frac{H}{c} \; \left( \nu \frac{\partial I_\nu}{\partial \nu} -
  3 I_\nu \right) = -\kappa_\nu I_\nu + j_\nu .
\end{equation}
%
Here $I_\nu \equiv I(\nu, \mathbf{x}, \Omega, t)$ is the radiation
specific intensity in units of energy per time $t$ per solid angle per
unit area per frequency $\nu$.  $H = \dot{a}/a$ is the Hubble
constant, where $a$ is the scale factor.  $\bar{a} = a/a_{em}$ is the
ratio of scale factors at the current time and time of emission.  The
second term represents the propagation of radiation, where the factor
$1/a$ accounts for cosmic expansion.  The third term describes both
the cosmological redshift and dilution of radiation.  On the right
hand side, the first term considers the absorption coefficient
$\kappa_\nu \equiv \kappa_\nu(\mathbf{x},\nu,t)$, and the second term
$j_\nu \equiv j_\nu(\mathbf{x},\nu,t)$ is the emission coefficient
that includes any point sources of radiation or diffuse radiation.

Solving this equation is difficult because of its high dimensionality;
however, we can make some appropriate approximations to reduce its
complexity in order to include radiation transport in numerical
calculations.  Typically timesteps in dynamic calculations are small
enough so that $\Delta a/a \ll 1$, therefore $\bar{a} = 1$ in any
given timestep, reducing the second term to $\hat{n} \partial
I_\nu/\partial \mathbf{x}$.  To determine the importance of the third
term, we evaluate the ratio of the third term to the second term.
This is $HL/c$, where $L$ is the simulation box length.  If this ratio
is $\ll 1$, we can ignore the third term.  For example at $z=5$, this
ratio is 0.1 when $L = c/H(z=5)$ = 53 proper Mpc.  In large boxes
where the light crossing time is comparable to the Hubble time, then
it could be important to consider cosmological redshifting and
dilution of the radiation, which we will describe later in
\S\ref{sec:ART}.  Thus equation (\ref{eqn:rteqn}) reduces to the
non-cosmological form in this local approximation,
%
\begin{equation}
  \frac{1}{c} \frac{\partial I_\nu}{\partial t} + 
  \hat{n} \frac{\partial I_\nu}{\partial \mathbf{x}} =
  -\kappa_\nu I_\nu + j_\nu .
\end{equation}
%
We choose to represent the source term $j_\nu$ as point sources of
radiation (e.g. stars, quasars) that emit radial rays that are
propagated along the direction $\hat{n}$.  Next we describe this
discretization and its contribution to the radiation field.

@@

The radiation field is calculated by integrating equation
\ref{eqn:rtray} along each ray, which is done by considering the
discretization of the ray into segments.  In the following section, we
assume the rays are monochromatic.  For convenience, we express the
integration in terms of optical depth $\tau = \int \kappa(r,t) \;
dr$, and for a ray segment,
%
\begin{equation}
  \label{eqn:dtau}
  d\tau = \sigma_{\rm abs}(\nu) n_{\rm abs} dr.
\end{equation}
Here $\sigma_{\rm abs}$ and $n_{\rm abs}$ are the cross section and
number density of the absorbing medium, respectively.  We use the
cell-centered density in our calculations but have experimented with
trilinearly interpolated densities \citep[see][]{Mellema06} without
producing considerable artifacts.  Equation (\ref{eqn:rtray}) has a
simple exponential analytic solution, and the photon flux of a ray is
reduced by
%
\begin{equation}
  \label{eqn:flux}
  dP = P \times (1 - e^{-\tau})
\end{equation}
as it crosses a cell.  We equate the photo-ionization rate to the
absorption rate, resulting in photon conservation \citep{Abel99_RT,
  Mellema06}.  Thus the photo-ionization $k_{\rm ph}$ and
photo-heating $\Gamma_{\rm ph}$ rates associated with a single ray are
%
\begin{equation}
  \label{eqn:kph}
  k_{\rm ph} = \frac{P (1 - e^{-\tau})}{n_{\rm abs} \; V_{\rm cell} \; dt_P},
\end{equation}
\begin{equation}
  \label{eqn:gamma}
  \Gamma_{\rm ph} = k_{\rm ph} \; (E_{\rm ph} - E_i),
\end{equation}
where $V_{\rm cell}$ is the cell volume, $E_{\rm ph}$ is the photon
energy, and $E_i$ is the ionization energy of the absorbing material.
In each cell, the photo-ionization and photo-heating rates from each
ray in the calculation are summed, and after the ray tracing is
complete, these rates can be used to update the ionization state and
energy of the cells.  Considering a system with only hydrogen
photo-ionizations and radiative recombinations, these changes are very
straightforward and is useful for illustrative purposes.  The change
in neutral hydrogen is
%
\begin{equation}
\frac{dn_{\rm H}}{dt} = \alpha_B n_e n_p - C_{\rm H} n_e n_{\rm
  H} - k_{\rm ph},
\end{equation}
where $\alpha_B = 2.59 \times 10^{-13} \mathrm{cm}^3 \mathrm{s}^{-1}$
is the recombination coefficient at 10$^4$ K in the Case B on-the-spot
approximation, in which all recombinations are locally reabsorbed,
\citep{Spitzer78}, and $C_{\rm H}$ is the collisional ionization rate.
However for more accurate solutions in calculations that consider
several chemical species, the photo-ionization rates are better
utilized in solvers that consider chemical networks
\citep[e.g.][]{Abel97}.

@@

Other radiative processes can also be important in some situations,
such as attenuation of radiation in the Lyman-Werner bands, secondary
ionizations from X-ray radiation, Compton heating of from scattered
photons, and radiation pressure.

@@

Molecular hydrogen can absorb photons in the Lyman-Werner bands that
are composed of 76 absorption lines ranging from 10.2 to 13.6 eV.
Each of these spectral lines can be modelled with a typical
exponential attenuation equation \citep{Ricotti01}, but
\citet{Draine96} showed that this self-shielding is well modeled with
the following relation to total \hh~column density
%
\begin{equation}
  \label{eqn:lwband}
  f_{\rm shield}(N_{\rm H2}) = \left\{ \begin{array}{l@{\quad}l}
      1 & (N_{\rm H2} \le 10^{14} \; \mathrm{cm}^{-2})\\
      (N_{\rm H2}/10^{14} \; \mathrm{cm}^{-2})^{-0.75} & (N_{\rm H2} >
        10^{14} \; \mathrm{cm}^{-2})
    \end{array} \right. .
\end{equation}
To incorporate this shielding function into the ray tracer, we store
the total \hh~column density and calculate the \hh~dissociation rate
by summing the contribution of all rays,
%
\begin{equation}
  \label{eqn:lwRT}
  k_{\rm diss} = \sum_{\rm rays} \frac{P \; \sigma_{\rm LW} \;
    \Omega_{\rm ray} \; r^2 \; dr}{A_{\rm cell} \; dV \; dt_P},
\end{equation}
where $\sigma_{\rm LW} = 3.71 \times 10^{18} \mathrm{cm}^2$ is the
effective cross-section of \hh~\citep{Abel97}.  To account for any
absorption in each photon package, we attenuate the photon number
flux by
%
\begin{equation}
  \label{eqn:LWdP}
  dP = P [f_{\rm shield}(N_{\rm H2} + dN_{\rm H2}) - f_{\rm shield}(N_{\rm H2})],
\end{equation}
where $dN_{\rm H2}$ is the \hh~column density in the current cell.

@@

Consider the solid angle $\Omega_{\rm ray}$ and photon flux $P$
associated with a single ray, and assume the flux is constant across
$\Omega_{\rm ray}$.  There exists a discrepancy between the geometry
cell face and HEALPix pixel when the pixel does not cover the entire
cell face, which is illustrated in Figure \ref{fig:covering}.  This
mismatch causes non-spherical artifacts and is most apparent in the
optically thin case, where the area of the pixel is dominant over $(1
- e^\tau)$ when calculating $k_{\rm ph}$.  One can avoid these
artifacts by increasing the sampling $\Phi_c$ to high values,
e.g. $>10$, but we have formulated a simple geometric correction to
the calculation of the radiation field.  This correction is not unique
to the HEALPix formalism but can be applied to any type of
pixelization.

@@

When the medium is optically thin to the radiation, radiation is only
attenuated by geometric dilution in the local approximation to
Eq. (\ref{eqn:rteqn}), i.e. the inverse square law.  We can minimize
the computational work of ray tracing in the optically thin regime by
utilizing this fact.  Here we track the total column density $N_{\rm
  abs}$ and the equivalent total optical depth $\tau$ traversed by the
photon.  If $\tau < \tau_{\rm thin} \sim 0.1$ after the ray exits the
cell, we calculate the photo-ionization and photo-heating rates
directly from the incoming ray instead of the luminosity of the
source.  This should only be evaluated once per cell per radiation
source.
\begin{equation}
  \label{eqn:optthin}
  k_{\rm ph} = \frac{\sigma_{\rm abs} \; P}{dt_P \; \theta_{\rm pix}}
  \; \frac{r_{\rm cell}}{r_{\rm ray}}.
\end{equation}
Note that the photon number $P$ in the ray has already been
geometrically diluted by ray splitting.  Here $r_{\rm cell}$ and
$r_{\rm ray}$ are the radii from the radiation source to the cell
center and where the ray exits the cell.  Thus the last factor
corrects the flux to a value appropriate for the cell center.  The
photo-heating ray is calculated in the same manner as the general
case, $\Gamma_{\rm ph} = k_{\rm ph} (E_{\rm ph} - E_i)$.  No photons
are removed from the ray.  With this method, we only require one ray
travel through each cell where the gas is optically thin, thus
reducing the computational expense.

@@

\enzo~is a parallel block-structured AMR \citep{BergerAMR} code that
is publicly available \citep{BryanNorman1997, OShea2004}.  In this
section, we describe our parallel implementation of the adaptive ray
tracing method into \enzo.  First we explain the programming design of
handling the ``photon packages'' that are traced along the adaptive
rays.  We use the terms photon packages and rays interchangeably.  Next
we focus on the details of the radiation hydrodynamics and then the
importance of correct time-stepping.  Last we give our parallelization
strategy of tracing rays through an AMR hierarchy.  This
implementation is included in the \texttt{v2.0} public version of
\enzo.

@@

\subsubsection{Time averaged quantities within a timestep}
